<<<<<<< HEAD
---
title: "IODS course project"
author: "Niina Niinimäki"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# Introduction to Open Data Science - Course Project

```{r child = "chapter1.Rmd"}
# This will include Chapter 1 (that is updated in its own file) in the document.
```
  
***
  
```{r child = "chapter2.Rmd"}
# This will include Chapter 2 (that is updated in its own file) in the document.
```

***

(more chapters to be added similarly as we proceed with the course!)

=======
---
title: "IODS course project"
author: "Niina Niinimäki"
output:
  html_document:
    theme: cosmo
    toc: true
    toc_depth: 2
    fig_caption: true
    fig_width: 6
    fig_height: 4
---

***

# Introduction to Open Data Science - Course Project

```{r child = "chapter1.Rmd"}
# This will include Chapter 1 (that is updated in its own file) in the document.
```
  
***
  
```{r child = "chapter2.Rmd"}
# This will include Chapter 2 (that is updated in its own file) in the document.
```

***

(more chapters to be added similarly as we proceed with the course!)

>>>>>>> 7a8f9507f17be7902a23e693eb8718ff7869d871
